---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qacOutliers

<!-- badges: start -->
<!-- badges: end -->

The goal of qacOutliers is to detect multivariate and univariate outliers using seven different methods. This package consolidates methods from multiple different R packages and makes them more easily accessible. 

## Installation

You can install the development version of qacOutliers from [GitHub](https://github.com/chenning2011) with:

``` {r, message = F, results = F, warnings=F}
# install.packages("remotes")
remotes::install_github("https://github.com/chenning2011/qacOutliers/")
```

## How to use

To learn more about how to use this package, what outliers are, and the different functions available in this package, visit the [Getting Started](https://chenning2011.github.io/qacOutliers/articles/qacOutliers.html), [Multivariate](https://chenning2011.github.io/qacOutliers/articles/Multivariate.html), and [Univariate](https://chenning2011.github.io/qacOutliers/articles/Univariate.html) vignettes. You can find out more about how the specific functions work on the [Documentation](https://chenning2011.github.io/qacOutliers/reference/index.html) page. 


