---
title: "Univariate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Univariate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=F, message =F}
library(qacOutliers)
```

## What are univariate outliers? How do you detect them?

Univariate outliers are data points consisting of an extreme value for one variable. These values lie far from the dataset's central tendency. Here is a graphical example of a univariate outlier from the `Salaries` dataset in the `carData` package. The outlier is shown in red. 

```{r, echo = F, message = F}
library(ggplot2)
library(dplyr)
data(Salaries, package="carData")

#taking a subset of data with just the outlier so it can be red 
subset <- Salaries %>% 
  filter(salary > 220000)

#graph 
ggplot()+
  geom_histogram(data = Salaries, aes(x=salary), bins = 30)+
  geom_histogram(data = subset, aes(x=salary), fill = "red", bins = 30)+
  theme_minimal()+
  labs(title = "Distribution of Professor Salaries", x = "Salary (USD)", y = "Count")
```


While there are many different methods that can be used to detect univariate outliers, this package detects univariate outliers using three methods: creating a boxplot, using the Grubbs' test, and with the median absolute deviation test. 

The function detects outliers in the specified column(s) and provides the following output: 
(1) The value of the detected outliers and the corresponding row numbers.
(2) A graphical visualization showing the distribution of the data with the outliers highlighted according to the selected method.

## Boxplot Method 

The boxplot method for univariate outlier detection identifies extreme data points by examining the distribution of values using a boxplot. In this method, the "whiskers" of the boxplot represent the range of typical values within a dataset. In this function, outliers are defined as values that fall outside the whiskers, which are calculated as Q1−1.58×IQR for the lower bound and Q3+1.58×IQR for the upper bound, where Q1 and Q3 are the first and third quartiles, respectively, and IQR is the interquartile range (the difference between Q3 and Q1) to allow for a broader range and a different threshold for detecting extreme values. 1.58 is the length of the whiskers as multiple of IQR. 

This method helps quickly identify potential outliers, which are plotted individually and can indicate unusual or problematic data points that may warrant further investigation. 

```{r}
data(mtcars)
univOutliers(mtcars, "mpg", method="boxplot")
univOutliers(mtcars, "hp", method="boxplot")
univOutliers(mtcars, "qsec", method="boxplot")
```

The output will provide the identified outliers along with their corresponding row numbers, and a graphical representation will be generated with the outliers highlighted in red.

## Grubb's Test 

The Grubbs' Test is a simple technique that iteratively identifies outliers by testing the hypothesis that the largest or smallest value is an outlier. The Grubbs’ test requires the data to be normally distributed because it relies on mean and standard deviation. 

```{r}
data(mtcars)
univOutliers(mtcars, "mpg", method="grubbs")
univOutliers(mtcars, "hp", method="grubbs")
univOutliers(mtcars, "qsec", method="grubbs")
```

The output of Grubbs' test will provide the identified outliers and the corresponding row numbers. This method uses a Shapiro-Wilk test (shapiro.test) for normality. If the p-value of the normality test is less than 0.05, it issues a warning about normality because a high p-value could affect the accuracy of the Grubbs' Test.

The output of Grubbs' test will provide the identified outliers, the corresponding row numbers. This method uses a Shapiro-Wilk test (shapiro.test) for normality. If the p-value of the normality test is less than 0.05, it issues a warning about normality because a high p-value could affect the accuracy of the Grubbs' Test.

## MAD 

The MAD (Median Absolute Deviation) method for univariate outlier detection identifies outliers based on a robust measure of variability, the median absolute deviation. In this method, a red line represents the median of the values in the target vector. The upper CI limit is the median + 3MAD and the lower CI limit is median - 3MAD. The points outside of the confidence interval are the outliers detected by the MAD method.  

```{r}
data(mtcars)
univOutliers(mtcars, "mpg", method="mad")
univOutliers(mtcars, "hp", method="mad")
univOutliers(mtcars, "qsec", method="mad")
```

The output will provide the identified outliers along with their corresponding row numbers, and a graphical representation will be generated with the outliers highlighted in red.
 
